First Season MWA EoR Power Spectrum Results at Redshift 7

### 2016-06-27 ms.tex
JCP comments

\begin{abstract}
The Murchison Widefield Array (MWA) has collected hundreds of hours of Epoch of 
Reionization (EoR) data and now faces the challenge of overcoming foreground and 
systematic contamination to reduce the data to a cosmological measurement. We introduce 
several novel analysis techniques such as cable reflection calibration, high resolution 
gridding kernels, diffuse foreground model subtraction, and quality control methods. Each 
change to the analysis pipeline is tested against a two dimensional power spectrum figure 
of merit to demonstrate improvement. We incorporate the new techniques into a deep 
integration of 32 hours of MWA data. This data set is used to place a systematic-limited 
upper limit on the cosmological power spectrum of $\Delta^2 \leq 2.7 \times 10^4$ 
mK$^2$ at $k=0.27$ h~Mpc$^{-1}$ and $z=7.1$, consistent with other published limits, and a 
modest improvement (factor of 1.4) over previous MWA results. While our result is 
systematic limited, we identify several strategies to overcome these systematics in future 
analysis.
\end{abstract}
\keywords{cosmology: observations - cosmology: reionization}
\maketitle

@@

The 21 cm hyperfine transition from neutral hydrogen residing in the IGM during reionization 
offers another promising method to study the EoR. Not only is the 21cm signal a 
\emph{direct} probe of the IGM, but, due to the narrow width of the transition and the 
relationship between observed frequency and line-of-sight distance, it can be used to map 
the full three dimensional space of the epoch. (For reviews, see \citealt{Furlanetto:2006} 
and \citealt{Morales:2010}.) The first generation of instruments with primary science goals 
to detect the 21cm EoR signal have been built, including the GMRT (Giant Metrewave 
Radio Telescope, \citealt{Paciga:2013}), LOFAR (LOw Frequency Array\footnote{http://
www.lofar.org}, \citealt{Yatawatta:2013}), PAPER (Precision Array for Probing the Epoch of 
Reionization\footnote{http://eor.berkeley.edu}, \citealt{Parsons:2010}), and the MWA 
(Murchison Widefield Array\footnote{http://www.mwatelescope.org}, \citealt{Tingay:2013}). 
Due to relatively low signal to noise, these instruments aim for a statistical measurement of 
the EoR in the form of a cosmological power spectrum. Meanwhile, the second generation 
21cm EoR machines is on its way with the Hydrogen Epoch of Reionization Array (HERA
\footnote{http://reionization.org}, \citealt{DeBoer:2016}), which will further 
refine the power spectrum measurement in early build-out stages, but will ultimately be 
capable of imaging the ionized bubbles of reionization in its later stages 
\citep{Beardsley:2015,Malloy:2013}.

@@

This paper serves two purposes: to demonstrate several new analysis techniques and their 
impact on power spectrum estimation, and to present the first attempt at a deep integration 
power spectrum from the MWA. Using a three hour test set of data, we introduce several 
novel techniques including calibration of cable reflection contamination, high resolution 
gridding kernels, subtraction of a diffuse foreground model, and development of quality 
control methods which will be crucial for deeper integrations. We apply these new 
techniques to a deep integration of data from the first semester of MWA observations 
(August, 2013 -- November, 2013) on a single EoR field and redshift range, $6.2<z<7.5$. 
While not expecting to detect the EoR with less than a hundred hours of observations, this 
intermediate integration will serve to identify and diagnose systematics, allowing for 
improved analysis in future work.

The remainder of this paper is organized as follows. In Section~\ref{sec:MWA} we briefly 
describe the MWA instrument and the observations used in this work, in Section~
\ref{sec:analysis} we describe our analysis pipeline, in Section~\ref{sec:techniques} we 
describe several novel techniques in our analysis, in Section~\ref{sec:deep} we discuss our 
efforts to apply the analysis pipeline to a deep data integration and compare with a reference pipeline for robust estimation, and we discuss future work 
in Section~\ref{sec:discussion}. Throughout this paper we use a $\Lambda$CDM 
cosmology with $\Omega_m=0.73$, $\Omega_\Lambda=0.27$, and $h = 0.7$, consistent 
with WMAP seven year results \citep{Komatsu:2011}. All distances and wavenumbers are in 
comoving coordinates.

@@

While the 21 cm EoR signal is priority, the MWA is a general observatory serving several 
science programs beyond cosmology including Galactic and extragalactic surveys, time 
domain astrophysics, solar monitoring, and ionospheric science. The array was thus 
designed with these programs in consideration and the layout was optimized using a 
pseudo-random antenna placement algorithm \citep{Beardsley:2012} to obtain a well 
behaved point spread function while retaining a dense core for EoR sensitivity 
\citep{Beardsley:2013}.\footnote{A drone flyover view of the array layout is available on YouTube at \url{https://youtu.be/yDWdTUzTUMo}.}
A full description of the science capabilities of the array is found in
\citealt{Bowman:2013}.
The high imaging capability of the MWA enables us to calibrate and subtract foregrounds
based on sky models, leveraging the full capability of the observatory for the
EoR experiment. 
This strategy is in contrast to more pointed EoR experiments like PAPER, 
which utilize a highly redundant layout to enhance sensitivity. Redundant arrays can 
exploit symmetries of the instrument for quick calibration and analysis, but have poor 
point spread functions, making foregrounds more difficult to characterize.

@@

The data for this work includes two sets. First we use a set of 94 high band observations of 
the EoR0 field from August 23, 2013, which were taken early in telescope operations and 
found to be particularly well behaved. The MWA EoR collaboration has designated this set 
as a ``golden data set'', which is used to build analysis tools and compare early results. All 
techniques demonstrated here will use this golden data set. In Section~\ref{sec:deep} we 
will use a larger data set to move toward a deep integration. This data set consists of all 
EoR0 high band observations beginning August 23, 2013 (including the golden data set), 
and concluding when the field was no longer accessible for the season: when the sun was 
above the horizon for most or all of the field's transit. The final observation of this set was 
on November 29, 2013. The individual snapshots are 112 seconds of data at 0.5 second, 
40 kHz resolution. In total we analyze 2,780 snapshots, or about 86.5 hours of data.

@@

In the first calibration step we allow our complex gains to account for any direction-independent 
response on a per antenna, per frequency channel, per polarization basis. To 
define our gains in this way, we have made two simplifications. First we push all 
direction-dependence of the antenna response into the model of the primary beam. FHD is capable 
of using separate models for each antenna, though this has not yet been implemented for 
the MWA due to lack of individual models. The second simplification is to ignore any cross 
terms between our different axes, for example a mutual coupling between two antennas. In 
principle these terms would introduce baseline dependent gains, rather than antenna 
dependent. We have not yet seen evidence of these terms in MWA data, and so we currently 
neglect them in our solutions.

@@

The top panel of Figure~\ref{fig:cables} shows a 2D power spectrum using the gain model 
described so far. Three of the horizontal lines in the EoR window can be attributed to the 
coarse band gaps, as they reside at the harmonics expected for 1.28~MHz periodic 
sampling. But the sampling cannot account for the fourth line, highlighted by the arrow. The 
corresponding delay time of the line, $\tau \approx 1.23$~$\mu$s, corresponds almost 
exactly to twice the signal travel time through the 150 meter cables (with velocity 0.81c). We 
therefore introduced a reflection term into our gain model, which allows our restricted gains 
to account for the interference of the incident signal with a round-trip reflected signal in the 
cable. 
A similar approach was taken at lower frequencies in \citealt{Ewall-Wice:2016}.
The full gain model is expressed as
\begin{equation}\label{eq:cal2}
\hat{g}_i(\nu)=B_{\alpha}(\nu)\left(P_i(\nu)+R_i(\nu)\right),
\end{equation}
where the reflection term can be further decomposed as
\begin{equation}
R_i(\nu) = R_{i,0} e^{-2\pi i \tau_{i} \nu}.
\end{equation}
The reflection coefficient, $R_{i,0}$ is allowed to be complex, while the reflection delay, 
$\tau_i$ is real. After fitting for all other parameters, we fit for the reflection mode. This is 
done for antennas with suspected reflections, the most offensive of which is seen in the 
antennas with cables of 150 meters. Because the cables were not cut at exact lengths 
(variations on order tens of centimeters), we solve for both the reflection coefficient and 
delay by performing a direct Fourier transform to a highly over resolved delay grid and 
selecting the mode where the reflection amplitude is largest.

@@

We resolved this issue by increasing the resolution at which we form the kernel. While the 
most accurate answer is to model at infinite resolution, it is not computationally feasible to do 
so. Instead we chose a resolution at which the effect no longer materially 
impacts the power spectrum. With experimentation we found a kernel resolution of 0.007 
wavelengths was sufficient. The resulting model power spectrum is shown in the bottom 
panel of Figure~\ref{fig:beam_res}. The contamination within the EoR window now drops 
significantly below the predicted cosmological signal levels. This improvement in beam
model resolution came at the cost of 20\% increase in memory usage for our imaging
pipeline.

@@

\begin{figure}
\begin{center}
\subfigure[]{\includegraphics[width=\columnwidth]{window_x_no_cut_box_pointings.pdf}\label{fig:wedge_cut1}}
~
\subfigure[]{\includegraphics[width=\columnwidth]{window_ratio_cut_v1_box_pointings.pdf}\label{fig:wedge_cut2}}
\caption[Window power cut]{
\emph{Top:} Window power for each observation in our data set. The window power is 
calculated from the delay spectrum of uncalibrated data as a fast quality metric. Each color 
is a distinct day of observing, and the vertical dashed lines represent pointing shifts, with 
the pointing numbers indicated at the bottom of the plot above the horizontal axis. The black 
box indicates observations which passed this cut. \emph{Bottom:} Data cut based on 
window power polarization ratio. For each snapshot that passed the total power cut, we plot 
the ratio of window powers in the N-S and E-W polarizations. Clear outliers can be seen, 
including the whole of pointing +3. The black box again indicates the selected data.
\label{fig:wedge_cut}
}
\end{center}
\end{figure}

@@

The bandwidth of the MWA and our data set is 30.72 MHz, but in order to avoid the effects 
of cosmic evolution over the span of our measured redshift range, we limit observations to 
about 8 MHz, or $\Delta z \sim 0.4$ at our frequency. We do this by dividing the band into 
three overlapping sub-bands of 15.36 MHz each, which, after applying the Blackman-Harris 
window function, will result in effective bandwidths of 7.68 MHz. We label these sub-bands 
as ``low" ($z\approx 7.1$), ``mid" ($z\approx 6.8$), and ``high" ($z\approx 6.5$).

@@

The final mask we apply is the wedge. We found that a small buffer beyond the horizon was
necessary to completely mask out the wedge, consistent with \citealt{Dillon:2015}. We
implement the buffer by extending the horizon line by 15 degrees. This line is shown as the
dashed diagonal line in the left panel of Figure~\ref{fig:slices}.

@@

\begin{figure}
\includegraphics[width=\columnwidth]{1D_spectra_comparison.pdf}
\caption{
Here we compare the upper limits from the FHD$+$\eppsilon pipeline with those from the 
RTS$+$CHIPS pipeline. The magenta lines are repeated from Figure~\ref{fig:1d_ps}, and the gray lines
are the power plus two sigma thermal error bars lines generated by CHIPS. 
Due to the logarithmic binning used by CHIPS, some bins are empty, resulting in gaps in the
spectra.
CHIPS does not 
excise the coarse band harmonic lines, and so we see their strong effect where there are 
gaps in the magenta lines. The most notable differences are the ability of the RTS$+$CHIPS pipe
to access lower $k$ modes, while the FHD$+$\eppsilon pipe suppresses moe power before and
after the first coarse band harmonic. While the different analysis methods result in slightly
different upper limits, they are generally in agreement. 
\label{fig:ps_comparison}
}
\end{figure}

>>>>

\begin{abstract}
The Murchison Widefield Array (MWA) has collected hundreds of hours of Epoch of 
Reionization (EoR) data and now faces the challenge of overcoming foreground and 
systematic contamination to reduce the data to a cosmological measurement. We introduce 
several novel analysis techniques such as cable reflection calibration, hyper-resolution 
gridding kernels, diffuse foreground model subtraction, and quality control methods. Each 
change to the analysis pipeline is tested against a two dimensional power spectrum figure 
of merit to demonstrate improvement. We incorporate the new techniques into a deep 
integration of 32 hours of MWA data. This data set is used to place a systematic-limited 
upper limit on the cosmological power spectrum of $\Delta^2 \leq 2.7 \times 10^4$ 
mK$^2$ at $k=0.27$ h~Mpc$^{-1}$ and $z=7.1$, consistent with other published limits, and a 
modest improvement (factor of 1.4) over previous MWA results. While our result is 
systematic limited, we identify several strategies to overcome these systematics in future 
analysis.
\end{abstract}
\keywords{cosmology: observations - cosmology: reionization}
\maketitle

@@

The 21 cm hyperfine transition from neutral hydrogen residing in the IGM during reionization 
offers another promising method to study the EoR. Not only is the 21cm signal a 
\emph{direct} probe of the IGM, but, due to the narrow width of the transition and the 
relationship between observed frequency and line-of-sight distance, it can be used to map 
the full three dimensional space of the epoch. (For reviews, see \citealt{Furlanetto:2006}, 
\citealt{Morales:2010}, \citealt{Zaroubi:2013}, and \citealt{Pritchard:2012}) The first generation of instruments with primary science goals 
to detect the 21cm EoR signal have been built, including the GMRT (Giant Metrewave 
Radio Telescope, \citealt{Paciga:2013}), LOFAR (LOw Frequency Array\footnote{http://
www.lofar.org}, \citealt{vanHaarlem:2013, Yatawatta:2013}), PAPER (Precision Array for Probing the Epoch of 
Reionization\footnote{http://eor.berkeley.edu}, \citealt{Parsons:2010}), and the MWA 
(Murchison Widefield Array\footnote{http://www.mwatelescope.org}, \citealt{Tingay:2013, Bowman:2013}). 
Due to relatively low signal to noise, these instruments aim for a statistical measurement of 
the EoR in the form of a cosmological power spectrum. Meanwhile, the second generation 
21cm EoR machines is on its way with the Hydrogen Epoch of Reionization Array (HERA
\footnote{http://reionization.org}, \citealt{DeBoer:2016}), which will further 
refine the power spectrum measurement in early build-out stages, but will ultimately be 
capable of imaging the ionized bubbles of reionization in its later stages 
\citep{Beardsley:2015,Malloy:2013}.

@@

This paper serves two purposes: to demonstrate several new analysis techniques and their 
impact on power spectrum estimation, and to present the first deep integration 
power spectrum from the MWA. Using a three hour test set of data, we introduce several 
novel techniques including calibration of cable reflection contamination, high resolution 
gridding kernels, subtraction of a diffuse foreground model, and development of quality 
control methods which will be crucial for deeper integrations. We apply these new 
techniques to a deep integration of data from the first semester of MWA observations 
(August, 2013 -- November, 2013) on a single EoR field and redshift range, $6.2<z<7.5$. 
While not expecting to detect the EoR with less than a hundred hours of observations, this 
intermediate integration will serve to identify and diagnose systematics, allowing for 
improved analysis in future work.

The remainder of this paper is organized as follows. In Section~\ref{sec:MWA} we briefly 
describe the MWA instrument and the observations used in this work, in Section~
\ref{sec:analysis} we describe our analysis pipeline, in Section~\ref{sec:techniques} we 
describe several novel techniques in our analysis, in Section~\ref{sec:deep} we discuss our 
efforts to apply the analysis pipeline to a deep data integration and compare with an alternative analysis pipeline for robust estimation, and we discuss future work 
in Section~\ref{sec:discussion}. Throughout this paper we use a $\Lambda$CDM 
cosmology with $\Omega_m=0.73$, $\Omega_\Lambda=0.27$, and $h = 0.7$, consistent 
with WMAP seven year results \citep{Komatsu:2011}. All distances and wavenumbers are in 
comoving coordinates.

@@

While the 21 cm EoR signal is priority, the MWA is a general observatory serving several 
science programs beyond cosmology including Galactic and extragalactic surveys, time 
domain astrophysics, solar monitoring, and ionospheric science. The array was thus 
designed with these programs in consideration and the layout was optimized using a 
pseudo-random antenna placement algorithm \citep{Beardsley:2012} to obtain a well 
behaved point spread function while retaining a dense core for EoR sensitivity 
\citep{Beardsley:2013}.\footnote{A drone flyover view of the array layout is available on YouTube at \url{https://youtu.be/yDWdTUzTUMo}.}
A full description of the science capabilities of the array is found in
\citealt{Bowman:2013}.
The high imaging capability of the MWA enables us to calibrate and subtract foregrounds
based on sky models, leveraging the full capability of the observatory for the
EoR experiment. 
This strategy is in contrast to more targeted EoR experiments like PAPER, 
which utilize a highly redundant layout to enhance sensitivity. Redundant arrays can 
exploit symmetries of the instrument for quick calibration and analysis, but have poor 
point spread functions, making foregrounds more difficult to characterize.

@@

The data for this work includes two sets. First we use a set of 94 two-minute high band observations of 
the EoR0 field from August 23, 2013, which were taken early in telescope operations and 
found to be particularly well behaved. The MWA EoR collaboration has designated this set 
as a ``golden data set'', which is used to build analysis tools and compare early results. All 
techniques demonstrated here will use this golden data set \cite[e.g.][]{Jacobs:2016, Trott:2016, Pober:2016, Dillon:2015}. In Section~\ref{sec:deep} we 
will use a larger data set to move toward a deep integration. This data set consists of all 
EoR0 high band observations beginning August 23, 2013 (including the golden data set), 
and concluding when the field was no longer accessible for the season: when the sun was 
above the horizon for most or all of the field's transit. The final observation of this set was 
on November 29, 2013. The individual snapshots are 112 seconds of data at 0.5 second, 
40 kHz resolution. In total we analyze 2,780 snapshots, or about 86.5 hours of data.

@@

In the first calibration step we allow our complex gains to account for any direction-independent 
response on a per antenna, per frequency channel, per polarization basis. To 
define our gains in this way, we have made two simplifications. First we push all 
direction-dependence of the antenna response into the model of the primary beam. FHD is capable 
of using separate models for each antenna, though this has not been implemented in this 
work. The second simplification is to ignore any cross 
terms between our different axes, for example a mutual coupling between two antennas. In 
principle these terms would introduce baseline dependent gains, rather than antenna 
dependent. We have not yet seen evidence of these terms in MWA data, and so we currently 
neglect them in our solutions.

@@

The top panel of Figure~\ref{fig:cables} shows a 2D power spectrum using the gain model 
described so far. Three of the horizontal lines in the EoR window can be attributed to the 
coarse band gaps, as they reside at the harmonics expected for 1.28~MHz periodic 
sampling. But the sampling cannot account for the fourth line, highlighted by the arrow. The 
corresponding delay time of the line, $\tau \approx 1.23$~$\mu$s, corresponds almost 
exactly to twice the signal travel time through the 150 meter cables (with velocity 0.81c). We 
therefore introduced a reflection term into our gain model, which allows our restricted gains 
to account for the interference of the incident signal with a round-trip reflected signal in the 
cable. 
A similar approach was taken at lower frequencies in \citealt{Ewall-Wice:2016}.
The full gain model is expressed as
\begin{equation}\label{eq:cal2}
\hat{g}_i(\nu)=B_{\alpha}(\nu)\left(P_i(\nu)+R_i(\nu)\right),
\end{equation}
where the reflection term can be further decomposed as
\begin{equation}
R_i(\nu) = R_{i,0} e^{-2\pi i \tau_{i} \nu}.
\end{equation}
The reflection coefficient, $R_{i,0}$ is allowed to be complex, while the reflection delay, 
$\tau_i$ is real. After fitting for all other parameters, we fit for the reflection mode. This is 
done for antennas with suspected reflections, the most offensive of which is seen in the 
antennas with cables of 150 meters. Because the cables were not cut at exact lengths 
(variations on order tens of centimeters), we solve for both the reflection coefficient and 
delay by performing a direct Fourier transform to a highly over resolved delay grid and 
selecting the mode where the reflection amplitude is largest. 
This fitting produces a single reflection coefficient for each antenna with nominal cable
length of 150 meters.

@@

We resolved this issue by increasing the resolution at which we form the kernel. While the 
most accurate answer is to model at infinite resolution, it is not computationally feasible to do 
so. Instead we chose a resolution at which the effect no longer materially 
impacts the power spectrum. With experimentation we found a kernel resolution of 0.007 
wavelengths was sufficient. The resulting model power spectrum is shown in the bottom 
panel of Figure~\ref{fig:beam_res}. 
The contamination within the EoR window now drops significantly lower. Of course, without
knowing the exact level of the cosmological signal, we cannot know if this level is sufficiently 
low. This effect may need to be revisited if the EoR power spectrum is lower than 
predicted.
The improvement in beam
model resolution came at the cost of 20\% increase in memory usage for our imaging
pipeline.

@@

\begin{figure}
\begin{center}
\subfigure[]{\includegraphics[width=\columnwidth]{window_x_no_cut_box_pointings.pdf}\label{fig:wedge_cut1}}
~
\subfigure[]{\includegraphics[width=\columnwidth]{window_ratio_cut_v1_box_pointings.pdf}\label{fig:wedge_cut2}}
\caption[Window power cut]{
\emph{Top:} Window power for each observation in our data set. The window power is 
calculated from the delay spectrum of uncalibrated data as a fast quality metric. 
Because uncalibrated data is used, the units are arbitrary.
Each color 
is a distinct day of observing, and the vertical dashed lines represent pointing shifts, with 
the pointing numbers indicated at the bottom of the plot above the horizontal axis. The black 
box indicates observations which passed this cut. \emph{Bottom:} Data cut based on 
window power polarization ratio. For each snapshot that passed the total power cut, we plot 
the ratio of window powers in the N-S and E-W polarizations. Clear outliers can be seen, 
including the whole of pointing +3. The black box again indicates the selected data.
\label{fig:wedge_cut}
}
\end{center}
\end{figure}

@@

The bandwidth of the MWA and our data set is 30.72 MHz, but in order to avoid the effects 
of cosmic evolution over the span of our measured redshift range, we limit observations to 
about 8 MHz, or $\Delta z \sim 0.3$ at our frequency. We do this by dividing the band into 
three overlapping sub-bands of 15.36 MHz each, which, after applying the Blackman-Harris 
window function, will result in effective bandwidths of 7.68 MHz. We label these sub-bands 
as ``low" ($z\approx 7.1$), ``mid" ($z\approx 6.8$), and ``high" ($z\approx 6.5$).

@@

The final mask we apply is the wedge. We found that a small buffer beyond the horizon was
necessary to completely mask out the wedge, consistent with \citealt{Dillon:2015}. We
implement the buffer by increasing the slope of the horizon line by 14\%.
This line is shown as the
dashed diagonal line in the left panel of Figure~\ref{fig:slices}.

@@

\begin{figure}
\includegraphics[width=\columnwidth]{1D_spectra_comparison.pdf}
\caption{
Here we compare the upper limits from the FHD$+$\eppsilon pipeline with those from the 
RTS$+$CHIPS pipeline. The magenta lines are repeated from Figure~\ref{fig:1d_ps}, and the gray lines
are the power plus two sigma thermal error bars lines generated by CHIPS. 
Due to the logarithmic binning used by CHIPS, some bins are empty, resulting in gaps in the
spectra.
CHIPS does not 
excise the coarse band harmonic lines, and so we see their strong effect where there are 
gaps in the magenta lines. The most notable differences are the ability of the RTS$+$CHIPS pipe
to access lower $k$ modes, while the FHD$+$\eppsilon pipe suppresses more power before and
after the first coarse band harmonic. While the different analysis methods result in slightly
different upper limits, they are generally in agreement. 
\label{fig:ps_comparison}
}
\end{figure}
